---
title: "Confidence Intervals in linear regression"
description: |
  In this post I compare base graphics and ggplot for drawing confidence intervals in linear regression.
author:
  - name: "Tamas Schauer"
    url: https://github.com/tschauer
    affiliation: LMU - Biomedical Center
    affiliation_url: https://www.compbio.bmc.med.uni-muenchen.de/about_us/people/index.html
date: 2020-09-17
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
base_url: https://tschauer.github.io/blog/
twitter:
  site: "@tamas_schauer"
  card: summary_large_image
categories:
  - statistics
  - dataviz
repository_url: https://github.com/tschauer/blog/tree/gh-pages
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

---

## Setup

```{r}
set.seed(6)

n1 <- 10
n2 <- 50

x1 <- rnorm(n1, sd=2)
y1 <- 5 + 2*x1 + rnorm(n1, sd=1)

x2 <- rnorm(n2, sd=2)
y2 <- 10 + 1*x2 + rnorm(n2, sd=1)

x <- c(x1,x2)
y <- c(y1,y2)

df <- data.frame(x = x, 
                 y = y,
                 group = factor(c(rep("g1", n1), 
                                  rep("g2", n2))))

head(df)
```

## Linear regression

### Fit Summary

```{r}
fit1 <- lm(y1 ~ x1)
summary(fit1)
```

```{r}
fit2 <- lm(y2 ~ x2)
summary(fit2)
```

```{r}
fit <- lm(y ~ group + x + group:x, data = df)
summary(fit)
```



<!-- ### Diagnostics -->

<!-- ```{r, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'} -->
<!-- par(mfrow=c(2,2)) -->
<!-- plot(fit1, pch=20) -->
<!-- ``` -->

<!-- ```{r, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'} -->
<!-- par(mfrow=c(2,2)) -->
<!-- plot(fit2, pch=20) -->
<!-- ``` -->

<!-- ```{r, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'} -->
<!-- par(mfrow=c(2,2)) -->
<!-- plot(fit, pch=20) -->
<!-- ``` -->



### Confidence intervals

```{r}
pred1 <- data.frame(x1 = seq(min(x1),max(x1),0.1))
pred1 <- cbind(pred1, predict(object = fit1, newdata = pred1, interval = "confidence"))

head(pred1)
```

```{r}
pred2 <- data.frame(x2 = seq(min(x2), max(x2),0.1))
pred2 <- cbind(pred2, predict(object = fit2, newdata = pred2, interval = "confidence"))

head(pred2)
```

```{r}
pred <- data.frame(x = rep(seq(min(x),max(x),0.1), each=2), group = c("g1","g2"))
pred <- cbind(pred, predict(object = fit, newdata = pred, interval = "confidence"))

head(pred)
```

## Plot

### Base graphics

```{r, fig.align='center', fig.height=7, fig.width=12, dpi=200, out.width='80%'}
par(mfrow=c(1,2), mar = c(4,4,4,4), cex=1.25)

plot(y1 ~ x1, pch = 19, col = 1, 
     xlab = "x", ylab = "y",
     xlim = c(-5,6), ylim = c(-5,20))

points(y2 ~ x2, pch = 19, col = 2)

polygon(x = c(pred1$x1, rev(pred1$x1)), y = c(pred1$lwr, rev(pred1$upr)), col = "#DDDDDD88", border = NA)
polygon(x = c(pred2$x2, rev(pred2$x2)), y = c(pred2$lwr, rev(pred2$upr)), col = "#DDDDDD88", border = NA)

lines(fit ~ x1, data = pred1, col="#666666", lwd=2)
lines(fit ~ x2, data = pred2, col="#666666", lwd=2)


plot(y ~ x, data = df,
     pch = 19, col = df$group, 
     xlim = c(-5,6), ylim = c(-5,20))

pred_g1 <- pred[pred$group == "g1",]
pred_g2 <- pred[pred$group == "g2",]

polygon(x = c(pred_g1$x, rev(pred_g1$x)), y = c(pred_g1$lwr, rev(pred_g1$upr)), col = "#DDDDDD88", border = NA)
polygon(x = c(pred_g2$x, rev(pred_g2$x)), y = c(pred_g2$lwr, rev(pred_g2$upr)), col = "#DDDDDD88", border = NA)

lines(fit ~ x, data = pred_g1, col="#666666", lwd=2)
lines(fit ~ x, data = pred_g2, col="#666666", lwd=2)
```
* compare left and right: slopes are the same
* confidence intervals different
* it is beneficial to fit combined model (assumption should be checked though)

### ggplot

```{r, fig.align='center', fig.height=6, fig.width=12, dpi=200, out.width='80%'}
library(ggplot2)
library(ggpubr)

gg1 <- ggplot(df, aes(x=x, y=y, group = group)) + 
      theme_bw() +
      theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
            text = element_text(size=22)) +
      coord_cartesian(xlim = c(-5,6), ylim = c(-5,20)) + 
      geom_point(aes(colour = group), size = 2) +
      geom_smooth(method=lm, formula = y ~ x, se = TRUE,  aes(group = group), 
                  colour='#666666' , fill='#DDDDDD', alpha = 0.5) +
      scale_color_manual(values=c("black", "red"))


gg2 <- ggplot(df, aes(x=x, y=y, fill = group)) + 
      theme_bw() +
      theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
            text = element_text(size=22)) +
      coord_cartesian(xlim = c(-5,6), ylim = c(-5,20)) + 
      geom_point(aes(colour = group), size = 2) +
      geom_ribbon(data = pred, aes(x=x, y=fit,ymin = lwr, ymax = upr, group = group), 
                  fill='#DDDDDD', alpha = 0.5) +
      geom_line(data = pred, aes(y = fit), size = 1, colour="#666666") +
      scale_color_manual(values=c("black", "red"))

ggarrange(gg1, gg2,
          ncol = 2, nrow = 1)
```

* it looks the same as base graphics
* ```geom_smooth()``` is not the best choice because it fits 2 independent lines
* add confidence intervals calculated by ```predict()``` using ```geom_ribbon()```

