---
title: "R Base Graphics Tutorial"
description: |
  This tutorial helps to build R base graphics functions for plotting results of differential expression analysis by DESeq2. 
  The final functions are part of HelpersforDESeq2 package (under development).
author:
  - name: "Tamas Schauer"
    url: https://github.com/tschauer
    affiliation: LMU - Biomedical Center
    affiliation_url: https://www.compbio.bmc.med.uni-muenchen.de/about_us/people/index.html
date: 2020-05-15
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
base_url: https://tschauer.github.io/blog/
twitter:
  site: "@tamas_schauer"
  card: summary_large_image
categories:
  - dataviz
  - R base graphics
  - RNAseq
  - DESeq2
repository_url: https://github.com/tschauer/blog/tree/gh-pages
---



```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

---

## Setup

### Install Package

* The package is available on github. Installation is via devtools.
* Link: https://github.com/tschauer/HelpersforDESeq2


```{r eval=FALSE}
library(devtools)
install_github("tschauer/HelpersforDESeq2")
```

---

### Load Example Results

* Example tables can be found in the package extdata directory.

```{r}
library(HelpersforDESeq2)

data_dir <- system.file("extdata/", package = "HelpersforDESeq2")

res1 <- read.delim(paste0(data_dir, "res.HighMLEpATP-HighMLEmATP.txt"), row.names = 1, stringsAsFactors = F)
res2 <- read.delim(paste0(data_dir, "res.LowMLEpATP-LowMLEmATP.txt"), row.names = 1, stringsAsFactors = F)

# log10 transform baseMean
res1$log10baseMean <- log10(res1$baseMean+1)
res2$log10baseMean <- log10(res2$baseMean+1)

head(res1)
head(res2)
```

---

## Goal

* The goal is to shape the simplest scatterplot to almost publication quality. 

```{r message=FALSE, warning=FALSE,results=F,echo=F, fig.align='center', fig.height=3.75, fig.width=7, dpi=200, out.width='75%'}
par(mfrow=c(1,2), mar = c(4,4,2,2), oma = c(1,1,1,1), mgp = c(2,1,0))

plot(res1$log10baseMean,
     res1$log2FoldChange,
     main = "before")

plottingMA(res =  res1,
           main_title = "after",
           selection_ids = c("roX1","roX2"),
           selection_id_type = "gene_symbol",
           selection_point_size = 1,
           selection_text_label = TRUE,
           selection_shadow = FALSE,
           xlims = c(0, 6),
           ylims = c(-10,10),
           x_axis_by = 2,
           padj_cutoff = 0.01,
           show_legend = TRUE)
```

---

## Build Plot Step-by-Step

* simplest scatter plot

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(4,4,2,2), oma = c(4,4,4,4), mgp = c(2,1,0))

plot(res1$log10baseMean,
     res1$log2FoldChange)
```


* shape axes and points

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(4,4,2,2), oma = c(4,4,4,4), mgp = c(2,1,0))

plot(res1$log10baseMean,
     res1$log2FoldChange,
     xlab = "log10 mean counts",
     ylab = "log2 fold change",
     xlim = c(0,6),
     ylim = c(-10,10),
     col = rgb(0.7,0.7,0.7,0.5), pch=19, cex = 0.25)

abline(h=0, col="grey32")
```

* add point labels

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(4,4,2,2), oma = c(4,4,4,4), mgp = c(2,1,0))

plot(res1$log10baseMean,
     res1$log2FoldChange,
     xlab = "log10 mean counts",
     ylab = "log2 fold change",
     xlim = c(0,6),
     ylim = c(-10,10),
     col = rgb(0.7,0.7,0.7,0.5), pch=19, cex = 0.25)

abline(h=0, col="grey32")

selection_ids <- c("roX1","roX2")
selection_vector <- res1$gene_symbol %in% selection_ids
selection_color <- rgb(0.9,0.6,0,1)
        
points(res1$log10baseMean[selection_vector],
       res1$log2FoldChange[selection_vector],
       col = selection_color, pch=19, cex = 1.0)

legend("topright", legend =  "labeled", col = selection_color,
       bg = "white", border = NA, bty = "n", cex = 0.8, pch = 19)
```

* add point and text labels

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(4,4,2,2), oma = c(4,4,4,4), mgp = c(2,1,0))

plot(res1$log10baseMean,
     res1$log2FoldChange,
     xlab = "log10 mean counts",
     ylab = "log2 fold change",
     xlim = c(0,6),
     ylim = c(-10,10),
     col = rgb(0.7,0.7,0.7,0.5), pch=19, cex = 0.25)

abline(h=0, col="grey32")

selection_ids <- c("roX1","roX2")
selection_vector <- res1$gene_symbol %in% selection_ids
selection_color <- rgb(0.8,0,0,1)

points(res1$log10baseMean[selection_vector],
       res1$log2FoldChange[selection_vector],
       col = selection_color, pch=19, cex = 1.0)

text(res1$log10baseMean[selection_vector],
     res1$log2FoldChange[selection_vector], 
     res1$gene_symbol[selection_vector],
     col = selection_color, adj = c(0,-0.5))

legend("topright", legend =  "labeled", col = selection_color,
       bg = "white", border = NA, bty = "n", cex = 0.8, pch = 19)
```

* label all significant datapoints

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(4,4,2,2), oma = c(4,4,4,4), mgp = c(2,1,0))

plot(res1$log10baseMean,
     res1$log2FoldChange,
     xlab = "log10 mean counts",
     ylab = "log2 fold change",
     xlim = c(0,6),
     ylim = c(-10,10),
     col = rgb(0.7,0.7,0.7,0.5), pch=19, cex = 0.25)

abline(h=0, col="grey32")

selection_vector <- res1$padj < 0.01

points(res1$log10baseMean[selection_vector],
       res1$log2FoldChange[selection_vector],
       col = selection_color, pch=19, cex = 0.25)

legend("topright", legend =  "significant", col = selection_color,
       bg = "white", border = NA, bty = "n", cex = 0.8, pch = 19)
```

---

### Write a function

* reuse the code by writing a function

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=3.75, fig.width=7, dpi=200, out.width='75%'}
par(mfrow=c(1,2), mar = c(4,4,2,2), oma = c(1,1,1,1), mgp = c(2,1,0))

plottingMAbasics <- function(res,
                             padj_cutoff = 0.01){
        
        plot(res$log10baseMean,
             res$log2FoldChange,
             xlab = "log10 mean counts",
             ylab = "log2 fold change",
             xlim = c(0,6),
             ylim = c(-10,10),
             col = rgb(0.7,0.7,0.7,0.5), pch=19, cex = 0.25)
        
        abline(h=0, col="grey32")
        
        selection_vector <- res$padj < padj_cutoff
        
        points(res$log10baseMean[selection_vector],
               res$log2FoldChange[selection_vector],
               col = selection_color, pch=19, cex = 0.25)
        
        legend("topright", legend =  "significant", col = selection_color,
               bg = "white", border = NA, bty = "n", cex = 0.8, pch = 19) 
}

plottingMAbasics(res1)
plottingMAbasics(res2)

```

---

### Example Case of plottingMA Function

* Example of simple automatization using character string operations.
* Check how regular expressions work (*gsub*).
* Use *assign* and *get* to link names and objects.
* Iterate through names.
* This helps to avod mistakes by copy and paste.

```{r message=FALSE, warning=FALSE,results=T, fig.align='center', fig.height=3.75, fig.width=7, dpi=200, out.width='75%'}
par(mfrow=c(1,2), mar = c(4,4,2,2), oma = c(1,1,1,1), mgp = c(2,1,0))

res_file_names <- c("res.HighMLEpATP-Input.txt", "res.LowMLEpATP-Input.txt")

for(res_file_name in res_file_names){
        
        res_tmp <- read.delim(paste0(data_dir, res_file_name), row.names = 1, stringsAsFactors = F)
        
        res_name <- gsub(".txt","", res_file_name)
        assign(res_name, res_tmp)
        
}

res_names <- ls(pattern = "^res\\.")
res_names

for(res_name in res_names){
      
     plottingMA(res =  get(res_name),
           main_title = gsub("res.","",res_name),
           selection_ids = c("roX1","roX2"),
           selection_id_type = "gene_symbol",
           selection_point_size = 1,
           selection_text_label = TRUE,
           selection_shadow = FALSE,
           xlims = c(0, 6),
           ylims = c(-10,10),
           x_axis_by = 2,
           padj_cutoff = 0.01,
           show_legend = TRUE) 
}
```

---

### Log2FC - log2FC plot

* another common plot when exploring results

```{r message=FALSE, warning=FALSE,results=F, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(4,4,2,2), oma = c(4,4,4,4), mgp = c(2,1,0))

res1 <- get(res_names[1])
res2 <- get(res_names[2])
      
res_merged <- merge(res1, res2, by = "row.names")

plot(res_merged$log2FoldChange.x,
     res_merged$log2FoldChange.y,
     xlab = "log2FC res1",
     ylab = "log2FC res2",
     xlim = c(-6,6),
     ylim = c(-6,6),
     col = rgb(0.7,0.7,0.7,0.5), pch=19, cex = 0.25)

abline(h=0, v=0, col="grey32")
abline(coef = c(0,1), col="grey32", lty=2)

selection_ids = c("roX1","roX2")
selection_vector <- res_merged$gene_symbol.x %in% selection_ids
selection_color <- rgb(0.8,0,0,1)

points(res_merged$log2FoldChange.x[selection_vector],
       res_merged$log2FoldChange.y[selection_vector],
       col = selection_color, pch=19, cex = 1.0)

text(res_merged$log2FoldChange.x[selection_vector],
     res_merged$log2FoldChange.y[selection_vector], 
     res_merged$gene_symbol.x[selection_vector],
     col = selection_color, adj = c(0,-0.5))

legend("topleft", legend =  "labeled", col = selection_color,
       bg = "white", cex = 1, pch = 19)
```

---

### Example Case of plotLog2FC Function

```{r message=FALSE, warning=FALSE,results=T, fig.align='center', fig.height=6, fig.width=6, dpi=200, out.width='75%'}
par(mfrow=c(1,1), mar = c(5,5,1,1), oma = c(4,4,4,4), mgp = c(3,1,0))


plotLog2FC(res1 = get(res_names[1]),
           res2 = get(res_names[2]),
           main_title = "",
           x_label = paste("log2FC \n", gsub("res.","",res_names[1])),
           y_label = paste("log2FC \n", gsub("res.","",res_names[2])),
           lims = c(-6,6),
           point_size = 0.25,
           selection_ids = c("roX1","roX2"),
           selection_id_type = "gene_symbol",
           selection_point_size = 1,
           selection_legend = "labeled",
           selection_text_label = TRUE) 
```

---

### Check Function Code

* Link: https://github.com/tschauer/HelpersforDESeq2/tree/master/R

```{r}
plottingMA

plotLog2FC
```




